---
title: "Hausdorff Distance Matrix Computation"
subtitle: "Solutions"
author: "Aymeric Stamm"
date: today
format: html
toc: true
---

```{r setup}
#| include: false
mfdat <- readRDS("data/mfdat.rds")
dat <- readRDS("data/dat.rds")
dat <- dat[1:21]
library(purrr)
library(rlang)
library(roahd)
```

## Data & Goal

We have simulated 3D functional data for this lab that is provided in the Quarto
document in the `dat` object.

The `dat` object is a list of size $100$ containing $100$ three-dimensional
curves observed on a common grid of size $200$ of the interval $[0, 1]$.

As a result, each element of the `dat` list is a $3 \times 200$ matrix.

Here we focus on a subset of the data, the first $21$ curves, which looks like:
```{r data-viz}
plot(mfdat)
```

::: {.callout-tip icon=false}
## Objective

The goal is to implement a function similar to `stats::dist()` which computes
the pairwise distance matrix on this functional dataset using the Hausdorff
distance.
:::

## Hausdorff distance in R

We can implement the Hausdorff distance between two curves as:

```{r hausdorff_distance_vec}
hausdorff_distance_vec <- function(x, y) {
  P <- ncol(x)
  dX <- 1:P |>
    purrr::map_dbl(\(p) {
      min(colSums((y - x[, p])^2))
    }) |>
    max()
  dY <- 1:P |>
    purrr::map_dbl(\(p) {
      min(colSums((x - y[, p])^2))
    }) |>
    max()
  sqrt(max(dX, dY))
}
```

This version exploits the vectorized nature of R to compute the Hausdorff
distance via calls to `colSums()` and `max()`. Another version based on a double
loop is provided by the following `hausdorff_distance_for()` function:

```{r hausdorff_distance_for}
hausdorff_distance_for <- function(x, y) {
  P <- ncol(x)
  dX <- 0
  dY <- 0
  for (i in 1:P) {
    min_dist_x <- Inf
    min_dist_y <- Inf
    for (j in 1:P) {
      dist_x <- sum((y[, j] - x[, i])^2)
      if (dist_x < min_dist_x) {
        min_dist_x <- dist_x
      }
      dist_y <- sum((x[, j] - y[, i])^2)
      if (dist_y < min_dist_y) {
        min_dist_y <- dist_y
      }
    }
    if (min_dist_x > dX) {
      dX <- min_dist_x
    }
    if (min_dist_y > dY) {
      dY <- min_dist_y
    }
  }
  sqrt(max(dX, dY))
}
```

We can benchmark the two versions:

```{r}
#| echo: false
bm <- readRDS("data/hdist_r_bch.rds")
```

```{r}
#| eval: false
bm <- bench::mark(
  hausdorff_distance_vec(dat[[1]], dat[[2]]),
  hausdorff_distance_for(dat[[1]], dat[[2]])
)
```

```{r}
#| echo: false
bm |> 
  dplyr::select(expression, median, mem_alloc) |> 
  dplyr::mutate(
    expression = purrr::map_chr(expression, deparse),
    median = as.numeric(median) * 1000,
    mem_alloc = as.numeric(mem_alloc)
  ) |>
  gt::gt() |>
  gt::cols_label(
    expression = gt::md("**Expression**"),
    median = gt::md("**Median computation time**"),
    mem_alloc = gt::md("**Memory allocation**")
  ) |>
  gt::cols_align(align =  "left", columns = expression) |>
  gt::cols_align(align = "right", columns = median:mem_alloc) |>
  gt::cols_align_decimal() |>
  gt::fmt_number(columns = "median", decimals = 2, pattern = "{x} ms") |>
  gt::fmt_bytes(columns = "mem_alloc")
```

We conclude that the vectorized version is faster but has a huge memory
footprint compared to the loop-based version. This means that the vectorized
version is not suitable for even moderately large data sets.

## Pairwise distance matrix in R

::: {.callout-tip}
## dist objects

Take a look at the documentation of the `stats::dist()` function to understand
how to make an object of class `dist`.
:::

We can exploit the previous functions to compute the pairwise distance matrix
using the Hausdorff distance:

```{r dist_r_v1}
dist_r_v1 <- function(x, vectorized = FALSE) {
  hausdorff_distance <- if (vectorized) 
    hausdorff_distance_vec
  else 
    hausdorff_distance_for
  N <- length(x)
  out <- 1:(N - 1) |>
    purrr::map(\(i) {
      purrr::map_dbl((i + 1):N, \(j) {
        hausdorff_distance(x[[i]], x[[j]])
      })
    }) |>
    purrr::list_c()

  attributes(out) <- NULL
  attr(out, "Size") <- N
  lbls <- names(x)
  attr(out, "Labels") <- if (is.null(lbls)) 1:N else lbls
  attr(out, "Diag") <- FALSE
  attr(out, "Upper") <- FALSE
  attr(out, "method") <- "hausdorff"
  class(out) <- "dist"
  out
}
```

We can benchmark the two versions:

```{r}
#| echo: false
bm <- readRDS("data/dist_r_v1_bch.rds")
```

```{r}
#| eval: false
bm <- bench::mark(
  dist_r_v1(dat, vectorized = TRUE),
  dist_r_v1(dat, vectorized = FALSE)
)
```

```{r}
#| echo: false
bm |> 
  dplyr::select(expression, median, mem_alloc) |> 
  dplyr::mutate(
    expression = purrr::map_chr(expression, deparse),
    median = as.numeric(median),
    mem_alloc = as.numeric(mem_alloc)
  ) |>
  gt::gt() |>
  gt::cols_label(
    expression = gt::md("**Expression**"),
    median = gt::md("**Median computation time**"),
    mem_alloc = gt::md("**Memory allocation**")
  ) |>
  gt::cols_align(align =  "left", columns = expression) |>
  gt::cols_align(align = "right", columns = median:mem_alloc) |>
  gt::cols_align_decimal() |>
  gt::fmt_number(columns = "median", decimals = 2, pattern = "{x} s") |>
  gt::fmt_bytes(columns = "mem_alloc")
```

::: {.callout-tip}
## Memory footprint

We confirm that the vectorized version is not scalable to large datasets. Using
it on the full dataset actually requires 12GB of memory! We will therefore focus
on the loop-based version from now on.
:::

## futureverse

### Parallelize outer loop

```{r dist_r_v2}
dist_r_v2 <- function(x) {
  N <- length(x)
  out <- 1:(N - 1) |>
    furrr::future_map(\(i) {
      purrr::map_dbl((i + 1):N, \(j) {
        hausdorff_distance_for(x[[i]], x[[j]])
      })
    }) |>
    purrr::list_c()

  attributes(out) <- NULL
  attr(out, "Size") <- N
  lbls <- names(x)
  attr(out, "Labels") <- if (is.null(lbls)) 1:N else lbls
  attr(out, "Diag") <- FALSE
  attr(out, "Upper") <- FALSE
  attr(out, "method") <- "hausdorff"
  class(out) <- "dist"
  out
}
```

### Tweaking the chunk size

```{r dist_r_v3}
dist_r_v3 <- function(x) {
  N <- length(x)
  out <- 1:(N - 1) |>
    furrr::future_map(\(i) {
      purrr::map_dbl((i + 1):N, \(j) {
        hausdorff_distance_for(x[[i]], x[[j]])
      })
    }, .options = furrr::furrr_options(chunk_size = 1)) |>
    purrr::list_c()

  attributes(out) <- NULL
  attr(out, "Size") <- N
  lbls <- names(x)
  attr(out, "Labels") <- if (is.null(lbls)) 1:N else lbls
  attr(out, "Diag") <- FALSE
  attr(out, "Upper") <- FALSE
  attr(out, "method") <- "hausdorff"
  class(out) <- "dist"
  out
}
```

### Nested plan

```{r dist_r_v4}
dist_r_v4 <- function(x) {
  N <- length(x)
  out <- 1:(N - 1) |>
    furrr::future_map(\(i) {
      furrr::future_map_dbl((i + 1):N, \(j) {
        hausdorff_distance_for(x[[i]], x[[j]])
      })
    }, .options = furrr::furrr_options(chunk_size = 1)) |>
    purrr::list_c()

  attributes(out) <- NULL
  attr(out, "Size") <- N
  lbls <- names(x)
  attr(out, "Labels") <- if (is.null(lbls)) 1:N else lbls
  attr(out, "Diag") <- FALSE
  attr(out, "Upper") <- FALSE
  attr(out, "method") <- "hausdorff"
  class(out) <- "dist"
  out
}
```

### Convert to single loop

```{r dist_r_v5}
dist_r_v5 <- function(x) {
  N <- length(x)
  K <- N * (N - 1) / 2
  out <- furrr::future_map_dbl(1:K, \(k) {
    k <- k - 1
    i <- N - 2 - floor(sqrt(-8 * k + 4 * N * (N - 1) - 7) / 2.0 - 0.5);
    j <- k + i + 1 - N * (N - 1) / 2 + (N - i) * ((N - i) - 1) / 2;
    i <- i + 1
    j <- j + 1
    hausdorff_distance_for(x[[i]], x[[j]])
  })
  attributes(out) <- NULL
  attr(out, "Size") <- N
  lbls <- names(x)
  attr(out, "Labels") <- if (is.null(lbls)) 1:N else lbls
  attr(out, "Diag") <- FALSE
  attr(out, "Upper") <- FALSE
  attr(out, "method") <- "hausdorff"
  class(out) <- "dist"
  out
}
```

## C++ implementation

The fact that the `dist` function takes a list of matrices as input can be
handled by [{Rcpp}](https://rcpp.org/). However, there is no threadsafe wrapper
for lists due to the fact that they can store objects of different types and
thus cannot be serialized.

We therefore use a vector representation for an $L$-dimensional curve observed
on a grid of $P$ points. The vector is of length $L \times N$. This allows the
entire data set to be passed as a single matrix $X$ to the C++ function where
the $i$-*th* row reads:

$$
x_i^{(1)}(t_1), \dots, x_i^{(1)}(t_P), x_i^{(2)}(t_1), \dots, x_i^{(2)}(t_P), \dots, x_i^{(L)}(t_1), \dots, x_i^{(L)}(t_P).
$$

```{Rcpp}
#| file: src/hausdorff.cpp
```

The `hausdorff_distance_cpp()` function implements the Hausdorff distance
between two curves, where a curve is stored as a vector as described in the
beginning of the section. The function therefore takes an additional optional
argument to specify the dimension of the curve. The function has two
implementations with different input types (`Rcpp::NumericVector` and
`RcppParallel::RMatrix<double>::Row`). The latter is used to parallelize the
computation using the thread-safe accessor to vectors. Only the former is
exported to R.

The `listToMatrix()` function is used to convert a list of matrices to a single
matrix that stores the sample of curves in a format that can be passed to the
`dist_cpp_omp()` and `dist_rcppparallel()` functions in a thread-safe manner.

The `dist_cpp_omp()` and `dist_rcppparallel()` functions compute the Hausdorff
distance between all pairs of curves in the sample. The former uses
[OpenMP](https://www.openmp.org) to parallelize the computation, while the
latter uses [{RcppParallel}](https://rcppcore.github.io/RcppParallel/). Both
functions have two implementations with different input types
(`Rcpp::NumericMatrix` and `Rcpp::List`). The former can be wrapped with the
tread-safe accessor to matrices, while the latter is the one that is exported to
R and internally uses the `listToMatrix()` function to convert the input list to
a matrix.

The `dist_rcppparallel()` function parallelizes the computations via the
`RcppParallel::parallelFor()` function, which requires a `RcppParallel::Worker`
object. The `HausdorffDistanceComputer` class inherits from the
`RcppParallel::Worker` class and is used to implement exactly what a single
worker should do.

## Benchmark

```{r}
#| echo: false
bm <- readRDS("data/dist_all_bch.rds")
```

```{r}
#| eval: false
library(future)

bm <- bench::mark(
  sequential = dist_r_v1(dat, vectorized = FALSE),
  outer = {
    plan(multisession, workers = 4L)
    out <- dist_r_v2(dat)
    plan(sequential)
    out
  },
  chunksize = {
    plan(multisession, workers = 4L)
    out <- dist_r_v3(dat)
    plan(sequential)
    out
  },
  nested = {
    plan(list(
      tweak(multisession, workers = 2L),
      tweak(multisession, workers = I(2L))
    ))
    out <- dist_r_v4(dat)
    plan(sequential)
    out
  },
  singleloop = {
    plan(multisession, workers = 4L)
    out <- dist_r_v5(dat)
    plan(sequential)
    out
  },
  openmp1 = dist_cpp_omp(dat, dimension = 3L, ncores = 1L),
  openmp4 = dist_cpp_omp(dat, dimension = 3L, ncores = 4L),
  rcppparallel1 = dist_rcppparallel(dat, dimension = 3L, ncores = 1L),
  rcppparallel4 = dist_rcppparallel(dat, dimension = 3L, ncores = 4L)
)
```

```{r}
#| echo: false
bm |> 
  dplyr::select(expression, median, mem_alloc) |> 
  dplyr::mutate(
    lang = c(rep("R", 5), rep("C++", 4)),
    expression = names(expression),
    median = as.numeric(median) * 1000,
    mem_alloc = as.numeric(mem_alloc)
  ) |>
  dplyr::select(lang, expression, median, mem_alloc) |>
  gt::gt() |>
  gt::cols_label(
    lang = gt::md("**Language**"),
    expression = gt::md("**Expression**"),
    median = gt::md("**Median computation time**"),
    mem_alloc = gt::md("**Memory allocation**")
  ) |>
  gt::cols_align(align =  "left", columns = lang:expression) |>
  gt::cols_align(align = "right", columns = median:mem_alloc) |>
  gt::cols_align_decimal(columns = median:mem_alloc) |>
  gt::fmt_number(columns = "median", decimals = 2, pattern = "{x} ms") |>
  gt::fmt_bytes(columns = "mem_alloc") |> 
  gt::tab_header(
    title = "Comparison of different implementations",
    subtitle = "The computation time is given in milliseconds and the memory allocation in bytes."
  ) |> 
  # Add line color based on language
  gt::tab_style(
    style = gt::cell_fill(color = "lightskyblue1"),
    locations = gt::cells_body(rows = lang == "R")
  ) |>
  gt::tab_style(
    style = gt::cell_fill(color = "lightgoldenrod1"),
    locations = gt::cells_body(rows = lang == "C++")
  ) |> 
  # Color the header
  gt::tab_style(
    style = gt::cell_fill(color = "lightgray"),
    locations = gt::cells_column_labels()
  )
```

## Interpretation

**General observation.** Using C++ is much faster than using R. This is in
particular due to never copying the data. It also provides linear speed-up as
expected.

**R implementation.** The original Hausdorff distance implementation has a
double loop.

- The outer loop generates unbalanced chunks, which is not optimal (`outer`
entry).
- Load balancing is better with a fixed chunk size of 1 (`chunksize` entry). 
- Using nested parallelization does not help (`nested` entry) achieves
intermediate results. It is better than the outer loop but worse than the fixed
chunk size.
- Transforming the original double loop into a single loop (`singleloop` entry)
is as fast as parallelizing the outer loop using a chunk size of 1 (`singleloop`
entry) but its memory allocation is much lower. This is because, when the chunk
size is 1, a future is created for each iteration of the loop, and the whole
data is copied for each future. In contrast, the single loop version submits a
single future per worker which creates less copies of the data.
- In general, playing with chunk size and nested parallelization generates more
future objects and copies the data more often, which increases memory allocation.
